---
title: "702 PS-5"
author: "Sagnik Roy"
date: "2026-02-25"
output: word_document
---

## Problem Set - 3

### 3. Problem to demonstrate the role of qualitative (ordinal) predictors in addition to quantitative predictors in multiple linear regression

Consider “diamonds” data set in R. It is in the ggplot2 package. Make a list of all the ordinal categorical variables. Identify the response.

#### (a) Run a linear regression of the response on the quality of cut. Write the fitted regression model.

```{r,message=FALSE,warning=FALSE}
library(ggplot2)
str(diamonds)
grade=as.numeric(diamonds$cut)
f=function(x){
  if(x-5==0){return(c(0,0,0,0))
  }else if(x-5==-1){return(c(1,0,0,0))
  }else if(x-5==-2){return(c(1,1,0,0))
  }else if(x-5==-3){return(c(1,1,1,0))    
  }else{return(c(1,1,1,1))}}
p=t(sapply(grade,f))
colnames(p)=c("Premium","VGood","Good","Fair")
diamonds=cbind(diamonds,p)
summary(lm(price~Premium+VGood+Good+Fair,data=diamonds))
```

#### (b) Test whether the expected price of diamond with premium cut is significantly different from that of the ideal cut.

We will perform the following test:

$H_0: \beta_{\text{Premium}} - \beta_{\text{Ideal}} = 0$
***ag***
$H_1: \beta_{\text{Premium}} - \beta_{\text{Ideal}} \neq 0$

The test statistic under $H_0$ is given as: 

$t = \frac{\hat{\beta}_{\text{Premium}}}{SE(\hat{\beta}_{\text{Premium}})}$

The p-value we observe for the test is negligible (almost equal to 0 << $\alpha = 0.05$). Therefore, we conclude that there is a statistically significant difference between the two levels.

#### (c) What is the expected price of a diamond of ideal cut? 

The expected price of a diamond of ideal cut is $3457.54.

#### (d) Modify the regression model in (a) by incorporating the predictor “table”. Write the fitted regression model.

```{r}
summary(lm(price~Premium+VGood+Good+Fair+table,data=diamonds))
```

#### (e) Test for the significance of “table” in predicting the price of diamond.

For this we will be performing the student's t test.

We will perform the following test:

$H_0: \beta_{table} = 0$
***ag***
$H_1: \beta_{table} \neq 0$

The test statistic under $H_0$ is given as: 

t = $\frac{\hat\beta_{table}}{SE(\beta_{table})}$

The p-value we observe for the test is negligible (almost equal to 0 << $\alpha = 0.05$). Therefore, we conclude that table is a statistically significant predictor.

#### (f) Find the average estimated price of a diamond with an average table value and which is of fair cut.

The average estimated price of a diamond with an average table value and which is of fair cut is $4092.76

---

## Problem Set - 5

### 1. Problem to demonstrate the utility of K nearest neighbour regression over least squares regression.

Consider a setting with n = 1000 observations. Generate:

 (i) $x_{1i}$ from $N(0, 2^2)$ and $x_{2i}$ from $Poisson(λ = 1.5)$.
 
 (ii) $ε_i$ from $N(0, 1)$.
 
 (iii) $yi = −2 + 1.4x_{1i} − 2.6x_{2i} + ε_i$
 
We are using the seed as 123.

Split the data into train and test sets. Keep the first 800 observations as training data and the remaining as test data. Work out the following:

```{r}
set.seed(123)
n=1000
x1=rnorm(n,0,2);x2=rpois(n,1.5);e=rnorm(n)
y=-2+1.4*x1-2.6*x2+e
d=data.frame(y,x1,x2)
head(d)
gg=c(1:800)
train=d[gg,];test=d[-gg,]
```

#### 1. Fit a multiple linear regression equation of y on x1 and x2. Calculate test MSE.

```{r}
m1=lm(y~x1+x2,data=train)
summary(m1)
yhat=predict(m1,newdata=test)
test.mse=mean((yhat-test[,1])^2)
test.mse
```

#### 2. Fit a KNN model with k = 1, 2, 5, 9, 15. Calculate test MSE for each choice of k.

```{r,message=FALSE,warning=FALSE}
library(ggplot2)
library(lattice)
library(caret)
mseks=c()
ks=c(1,2,5,9,15)
for(i in 1:5)
{kreg=knnreg(y~x1+x2,data=train,k=ks[i])
 ykreg=predict(kreg,test)
 mseks[i]=mean((test$y-ykreg)^2)}
output=data.frame(ks,mseks,rep(test.mse,each=5))
colnames(output)=c("Choices of k","Test MSE under KNN","Test MSE under Multiple Linear Reg")
output
```

Now we will be considering the function:

$y_i = \frac{1}{(−2 + 1.4x_{1i} − 2.6x_{2i} + 2.9 x^2_{1i})} + 3.1sin{x_{2i}} - 1.5x_{1i}x^2_{2i} + ε_i$

Work out the problems in (1) and (2). Compare and comment on the results.

```{r}
y1=1/(-2+1.4*x1-2.6*x2+2.9*(x1^2))+3.1*sin(x2)-1.5*x1*(x2^2)+e
d1=data.frame(y1,x1,x2)
head(d1)
train1=d1[gg,];test1=d1[-gg,]
m2=lm(y1~x1+x2,data=train1)
summary(m2)
y1test=predict(m2,newdata=test1)
testmse1=mean((y1test-test1[,1])^2);testmse1
msek1=c()
for(i in 1:5)
{kr1=knnreg(y1~x1+x2,data=train1,k=ks[i])
 y1_knn=predict(kr1,test1)
 msek1[i]=mean((test1[,1]-y1_knn)^2)}
output2=data.frame(ks,msek1,rep(testmse1,each=5))
colnames(output2)=c("Choices of k","Test MSE under KNN","Test MSE under Multiple Linear Reg")
output2
```

From the above comparison, we observe that the KNN method performs significantly better than multiple linear regression, as it produces a much lower test mean squared error. This is because KNN can capture non-linear relationships, whereas multiple linear regression assumes linearity and therefore performs poorly in this case. Although the minimum MSE occurs when k=1, the increase in MSE for larger values of k is relatively small, suggesting that choosing a moderately large value of k would still be appropriate.